Combinatorial Proofs and Decomposition Theorems for First-order Logic\vspace{-3ex}

### 2021-01-25 main.tex
fix the rule permutation part

  We have the following derivations:
  $$
  \vlderivation{
    \vlin{\cD}{}{\forall x.A}{
      \vlin{\equiv}{(x \notin fv(\forall x.A))}{(\forall x.A) \vlor (\forall x.A)}{
        \vlin{\wrD}{}{\forall x.((\forall x.A) \vlor A)}{
          \vlhy{\forall x.\forall x.A}}}}}
  $$.
  $$
  \vlderivation{
    \vlin{\cD}{}{(A \vlor B) \vlan (C \vlor D)}{
      \vlin{\fequ}{}{((A \vlor B) \vlan (C \vlor D)) \vlor ((A \vlor B) \vlan (C \vlor D))}{
      \vlin{\wrD}{}{((A \vlor B) \vlan (C \vlor D)) \vlor ((B \vlor A) \vlan (D \vlor C))}{
	\vlin{\wrD}{}{((A \vlor B) \vlan (C \vlor D)) \vlor ((B \vlor A) \vlan D))}{
	  \vlin{\wrD}{}{((A \vlor B) \vlan (C \vlor D)) \vlor (B \vlan D))}{
	    \vlin{\wrD}{}{((A \vlor B) \vlan C) \vlor (B \vlan D)}{
	      \vlhy{(A \vlan C) \vlor (B \vlan D)}}}}}}}}
  $$.
  $$
  \vlderivation{
    \vlin{\cD}{}{\exists x.(A \vlor B)}{
      \vlin{\fequ}{}{\exists x.(A \vlor B)) \vlor (\exists x.(A \vlor B))}{
      \vlin{\wrD}{}{(\exists x.(A \vlor B)) \vlor (\exists x.(B \vlor A))}{
	\vlin{\wrD}{}{(\exists x.(A \vlor B)) \vlor (\exists x.B)}{
	  \vlhy{(\exists x.A) \vlor (\exists x.B)}}}}}}
  $$.
  $$
  \vlderivation{
    \vlin{\cD}{}{\forall x.(A \vlor B)}{
      \vlin{\fequ}{}{(\forall x.(A \vlor B)) \vlor (\forall x.(A \vlor B))}{
      \vlin{\wrD}{}{(\forall x.(A \vlor B)) \vlor (\forall x.(B \vlor A))}{
	\vlin{\wrD}{}{(\forall x.(A \vlor B)) \vlor (\forall x.B)}{
	  \vlhy{(\forall x.A) \vlor (\forall x.B)}}}}}}
  $$.
  \juihsuan{done. If needed, we can introduce the notion of open deduction to
reduce the size of derivations...}\lutz{I was thinking about that, but (i) it is probably not worth the effort, as we wont have many derivation, and (ii) it is hard to define rectified derivations this way.}
\end{proof}

@@

\begin{proof}
  First, a derivation consisting of an instance of $\wrD$ followed by $\rr \in \set{\acD, \cfaD,
\me, \mfaD, \mexD}$ can be either replaced by a derivation consisting of $\wrD$
only or the instance of $\wrD$ can be permuted downwards. We show some cases
here, and leave the others to the reader.
  \begin{equation*}
  \vlderivation{
    \vlin{\mfaD}{}{\Scons{\forall x.(A \vlor B)}}{
      \vlin{\wrD}{}{\Scons{(\forall x.A) \vlor (\forall x.B)}}{
        \vlhy{\Scons{\forall x.A}}}}}
  \leadsto
  \vlderivation{
    \vlin{\wrD}{}{\Scons{\forall x.(A \vlor B)}}{
      \vlhy{\Scons{\forall x.A}}}}
  \end{equation*}
  \begin{equation*}
  \vlderivation{
    \vlin{\me}{}{\Scons{(A \vlor B) \vlan (C \vlor D)}}{
      \vlin{\wrD}{}{\Scons{(A \vlan C) \vlor (B \vlan D)}}{
        \vlhy{\Scons{A \vlan C}}}}}
  \leadsto
  \vlderivation{
    \vlin{\wrD}{}{\Scons{(A \vlor B) \vlan (C \vlor D)}}{
      \vlin{\wrD}{}{\Scons{(A \vlor B) \vlan C}}{
        \vlhy{\Scons{A \vlan C}}}}}
  \end{equation*}
  \begin{equation*}
  \vlderivation{
    \vlin{\acD}{}{\Scons{a}}{
      \vlin{\wrD}{}{\Scons{a \vlor a}}{
        \vlhy{\Scons{a}}}}}
  \leadsto
  \vlderivation{
    \vlhy{\Scons{a}}}
  \end{equation*}

  For $\rr_1 \in \set{\me, \mfaD, \mexD}$, $\rr_2 \in \set{\acD, \cfaD}$,
$\rr_1$ can be permuted upwards over $\rr_2$ (with some $\fequ$ inserted). The only non-trivial case is shown below:
  \begin{equation*}
  \vlderivation{
    \vlin{\mfaD}{}{\Scons{\forall x.(A \vlor B)}}{
      \vlin{\cfaD}{}{\Scons{(\forall x.A) \vlor (\forall x.B)}}{
        \vlhy{\Scons{(\forall x.\forall x.A) \vlor (\forall x.B)}}}}}
  \leadsto
  \vlderivation{
    \vlin{\mfaD}{}{\Scons{\forall x.(A \vlor B)}}{
      \vlin{\fequ}{}{\Scons{(\forall x.A) \vlor (\forall x.B)}}{
        \vlin{\mfaD}{}{\Scons{\forall x.(\forall x.A \vlor B)}}{
          \vlhy{\Scons{(\forall x.\forall x.A) \vlor (\forall x.B)}}}}}}
  \end{equation*}
  \todo{permutation with $\fequ$}

@@

Interaction between two rules:
\begin{itemize}
\item $\cfaD/\fequ$:

@@

@@

  \begin{equation*}
  \vlderivation{
    \vlin{\acD}{}{a \vlor B}{
      \vlin{\fequ}{}{(a \vlor a) vlor B}{
        \vlin{\wrD}{}{a \vlor (a \vlor B)}{
          \vlhy{a}}}}}
  \leadsto
  \vlderivation{
    \vlin{\wrD}{}{a \vlor B}{
      \vlhy{a}}}
  \end{equation*}

@@

  \begin{equation*}
  \vlderivation{
    \vlin{\acD}{}{\forall x.a}{
      \vlin{\fequ}{(x \notin fv(a))}{\forall x.(a \vlor a)}{
        \vlin{\wrD}{}{a \vlor (\forall x.a)}{
          \vlhy{a}}}}}
  \leadsto     
  \end{equation*}

@@

  \begin{equation*}
  \vlderivation{
    \vlin{\me}{}{\forall x.((A \vlor B) \vlan (C \vlor D))}{
      \vlin{\fequ}{x \notin fv(B \vlan D)}{\forall x.((A \vlan C) \vlor (B \vlan
D))}{
        \vlin{\wrD}{}{(B \vlan D) \vlor (\forall x.(A \vlan C))}{
          \vlhy{B \vlan D}}}}}
  \leadsto
  \end{equation*}

>>>>

  We have the following derivations:
  $$
  \vlderivation{
    \vlin{\cD}{}{\forall x.A}{
      \vlin{\fequ}{(x \notin fv(A))}{\forall x.(A \vlor A)}{
        \vlin{\wrD}{}{A \vlor (\forall x.A)}{
          \vlhy{A}}}}}
  $$.
  $$
  \vlderivation{
    \vlin{\cD}{}{\forall x.A}{
      \vlin{\equiv}{(x \notin fv(\forall x.A))}{(\forall x.A) \vlor (\forall x.A)}{
        \vlin{\wrD}{}{\forall x.((\forall x.A) \vlor A)}{
          \vlhy{\forall x.\forall x.A}}}}}
  $$.
  $$
  \vlderivation{
    \vlin{\cD}{}{(A \vlor B) \vlan (C \vlor D)}{
      \vlin{\fequ}{}{((A \vlor B) \vlan (C \vlor D)) \vlor ((A \vlor B) \vlan (C \vlor D))}{
      \vlin{\wrD}{}{((A \vlor B) \vlan (C \vlor D)) \vlor ((B \vlor A) \vlan (D \vlor C))}{
	\vlin{\wrD}{}{((A \vlor B) \vlan (C \vlor D)) \vlor ((B \vlor A) \vlan D))}{
	  \vlin{\wrD}{}{((A \vlor B) \vlan (C \vlor D)) \vlor (B \vlan D))}{
	    \vlin{\wrD}{}{((A \vlor B) \vlan C) \vlor (B \vlan D)}{
	      \vlhy{(A \vlan C) \vlor (B \vlan D)}}}}}}}}
  $$.
  $$
  \vlderivation{
    \vlin{\cD}{}{\exists x.(A \vlor B)}{
      \vlin{\fequ}{}{\exists x.(A \vlor B)) \vlor (\exists x.(A \vlor B))}{
      \vlin{\wrD}{}{(\exists x.(A \vlor B)) \vlor (\exists x.(B \vlor A))}{
	\vlin{\wrD}{}{(\exists x.(A \vlor B)) \vlor (\exists x.B)}{
	  \vlhy{(\exists x.A) \vlor (\exists x.B)}}}}}}
  $$.
  $$
  \vlderivation{
    \vlin{\cD}{}{\forall x.(A \vlor B)}{
      \vlin{\fequ}{}{(\forall x.(A \vlor B)) \vlor (\forall x.(A \vlor B))}{
      \vlin{\wrD}{}{(\forall x.(A \vlor B)) \vlor (\forall x.(B \vlor A))}{
	\vlin{\wrD}{}{(\forall x.(A \vlor B)) \vlor (\forall x.B)}{
	  \vlhy{(\forall x.A) \vlor (\forall x.B)}}}}}}
  $$.
  \juihsuan{done. If needed, we can introduce the notion of open deduction to
reduce the size of derivations...}\lutz{I was thinking about that, but (i) it is probably not worth the effort, as we wont have many derivation, and (ii) it is hard to define rectified derivations this way.}
\end{proof}

@@

  \todo{permutation with $\fequ$}

@@

We first study the interactions between two rules (only non-trivial cases are
presented here):
\begin{itemize}
\item $\rr_1/\rr_2$, where $\rr_1 \in \set{\wrD, \wfaD}$ and $\rr_2 \in
\set{\acD, \cfaD, \me, \mfaD, \mexD}$:

\begin{equation*}
\vlderivation{
  \vlin{\acD}{}{a}{
    \vlin{\wrD}{}{a \vlor a}{ 
      \vlhy{a}}}}
\leadsto
\vlderivation{
  \vlhy{a}}
\end{equation*}

\begin{equation*}
\vlderivation{
  \vlin{\me}{}{(A \vlor B) \vlan (C \vlor D)}{
    \vlin{\wrD}{}{(A \vlan C) \vlor (B \vlan D)}{
      \vlhy{A \vlan C}}}}
\leadsto 
\vlderivation{
  \vlin{\wrD}{}{(A \vlor B) \vlan (C \vlor D)}{
    \vlin{\wrD}{}{(A \vlor B) \vlan C}{
      \vlhy{A \vlan C}}}}
\end{equation*}

\begin{equation*}
\vlderivation{
  \vlin{\mfaD}{}{\forall x.(A \vlor B)}{
    \vlin{\wrD}{}{(\forall x.A) \vlor (\forall x.B)}{
      \vlhy{\forall x.A}}}}
\leadsto
\vlderivation{
  \vlin{\wrD}{}{\forall x.(A \vlor B)}{
    \vlhy{\forall x.A}}}
\end{equation*}

\begin{equation*}
\vlderivation{
  \vlin{\cfaD}{}{\forall x.A}{
    \vlin{\wfaD}{(x \notin fv(\forall x.A))}{\forall x.\forall x.A}{
      \vlhy{\forall x.A}}}}
\leadsto
\vlderivation{
  \vlhy{\forall x.A}}
\end{equation*}

\begin{equation*}
\vlderivation{
  \vlin{\mfaD}{}{\forall x.(A \vlor B)}{
    \vlin{\wfaD}{(x \notin fv(A))}{(\forall x.A) \vlor (\forall x.B)}{
      \vlhy{A \vlor (\forall x.B)}}}}
\leadsto
\vlderivation{
  \vlin{\fequ}{(x \notin fv(A))}{\forall x.(A \vlor B)}{
    \vlhy{A \vlor (\forall x.B)}}} 
\end{equation*}

\item $\rr_1/\rr_2$, where $\rr_1 \in \set{\acD, \cfaD}$ and $\rr_2 \in
\set{\me, \mfaD, \mexD}$: 
  \begin{equation*}
  \vlderivation{
    \vlin{\mfaD}{}{\Scons{\forall x.(A \vlor B)}}{
      \vlin{\cfaD}{}{\Scons{(\forall x.A) \vlor (\forall x.B)}}{
        \vlhy{\Scons{(\forall x.\forall x.A) \vlor (\forall x.B)}}}}}
  \leadsto
  \vlderivation{
    \vlin{\mfaD}{}{\Scons{\forall x.(A \vlor B)}}{
      \vlin{\fequ}{}{\Scons{(\forall x.A) \vlor (\forall x.B)}}{
        \vlin{\mfaD}{}{\Scons{\forall x.(\forall x.A \vlor B)}}{
          \vlhy{\Scons{(\forall x.\forall x.A) \vlor (\forall x.B)}}}}}}
  \end{equation*}

\item $\cfaD/\fequ$:

@@

\item $\wfaD/\fequ$:

In the following two cases, we assume $x \neq y$ (otherwise they are trivial).

\begin{equation*}
\vlderivation{
  \vlin{\fequ}{}{\forall y.\forall x.A}{
    \vlin{\wfaD}{(x \notin fv(\forall y.A))}{\forall x.\forall y.A}{
      \vlhy{\forall y.A}}}}
\leadsto
\vlderivation{
  \vlin{\wfaD}{(x \notin fv(A))}{\forall y.\forall x.A}{
    \vlhy{\forall y.A}}}
\end{equation*}

\begin{equation*}
\vlderivation{
  \vlin{\fequ}{}{\forall x.\forall y.A}{
    \vlin{\wfaD}{(x \notin fv(A))}{\forall y.\forall x.A}{
      \vlhy{\forall y.A}}}}
\leadsto
\vlderivation{
  \vlin{\wfaD}{(x \notin fv(\forall y.A))}{\forall x.\forall y.A}{
    \vlhy{\forall y.A}}}
\end{equation*}

\begin{equation*}
\vlderivation{
  \vlin{\fequ}{(x \notin fv(B))}{(\forall x.A) \vlor B}{
    \vlin{\wfaD}{(x \notin fv(A \vlor B))}{\forall x.(A \vlor B)}{
      \vlhy{A \vlor B}}}}
\leadsto
\vlderivation{
  \vlin{\wfaD}{(x \notin fv(A))}{(\forall x.A) \vlor B}{
    \vlhy{A \vlor B}}}
\end{equation*}

\begin{equation*}
\vlderivation{
  \vlin{\fequ}{(x \notin fv(B))}{\forall x.(A \vlor B)}{
    \vlin{\wfaD}{(x \notin fv(A))}{(\forall x.A) \vlor B}{
      \vlhy{A \vlor B}}}}
\leadsto
\vlderivation{
  \vlin{\wfaD}{(x \notin fv(A \vlor B))}{\forall x.(A \vlor B)}{
    \vlhy{A \vlor B}}}
\end{equation*}

@@

  \begin{equation*}
  \vlderivation{
    \vlin{\acD}{}{a \vlor B}{
      \vlin{\fequ}{}{(a \vlor a) \vlor B}{
        \vlin{\wrD}{}{a \vlor (a \vlor B)}{
          \vlhy{a}}}}}
  \leadsto
  \vlderivation{
    \vlin{\wrD}{}{a \vlor B}{
      \vlhy{a}}}
  \end{equation*}

@@

  \begin{equation*}
  \vlderivation{
    \vlin{\acD}{}{\forall x.a}{
      \vlin{\fequ}{(x \notin fv(a))}{\forall x.(a \vlor a)}{
        \vlin{\wrD}{}{a \vlor (\forall x.a)}{
          \vlhy{a}}}}}
  \leadsto
  \vlderivation{
    \vlin{\wfaD}{(x \notin fv(a))}{\forall x.a}{
      \vlhy{a}}}
  \end{equation*}

@@

  \begin{equation*}
  \vlderivation{
    \vlin{\me}{}{\forall x.((A \vlor B) \vlan (C \vlor D))}{
      \vlin{\fequ}{(x \notin fv(B \vlan D))}{\forall x.((A \vlan C) \vlor (B \vlan
D))}{
        \vlin{\wrD}{}{(B \vlan D) \vlor (\forall x.(A \vlan C))}{
          \vlhy{B \vlan D}}}}}
  \leadsto
  \vlderivation{
    \vlin{\fequ}{}{\forall x.((A \vlor B) \vlan (C \vlor D))}{
      \vlin{\wrD}{}{\forall x.((B \vlor A) \vlan (D \vlor C))}{
        \vlin{\wrD}{}{\forall x.((B \vlor A) \vlan D)}{
          \vlin{\faD}{(x \notin fv(B \vlan D))}{\forall x.(B \vlan D)}{
            \vlhy{B \vlan D}}}}}}
  \end{equation*}
